Multiple imputation of incomplete multilevel data using Heckman selection models

### 2022-11-22 1.Paper/Paper.Rmd
update text

corres: "* Johanna Munoz. \\email{j.munozavila@umcutrecht.nl}"
presentaddress: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Universiteitsweg 100, 3584 CG Utrecht, the Netherlands
authormark: Uthor \emph{et al}.
articletype: Research article
received: 2022-11-01
revised: 2022-12-01
accepted: 2023-01-01
abstract: "Missing data is a common problem in medical research, and is commonly addressed using multiple imputation. Although traditional imputation methods allow for valid statistical inference when data are missing at random (MAR), their implementation is not justified when observations are clustered (e.g., within studies) or when the presence of missingness depends on unobserved information, i.e. the data are missing not at random (MNAR). Unfortunately, this situation is increasingly common, and typically arises when individual participant data (IPD) from multiple studies are combined. While several imputation methods have been proposed for addressing individual studies when data are MNAR, their application and validity in large datasets with clustering remains unclear. We therefore explored the consequence of MNAR data in IPD meta-analysis in-depth, and proposed a novel multilevel imputation method for common missing patterns in clustered datasets. This method is based on the principles of Heckman selection models and adopts a two-stage meta-analysis approach to impute binary and continuous variables that may be outcomes or predictors and that are systematically or sporadically missing. After evaluating the proposed imputation model in simulated scenarios, we illustrate it use in a cross-sectional community survey to estimate the prevalence of malaria parasitemia in children aged 2-10 years in five subregions in Uganda."

@@

Selection bias occurs when the inclusion of a subject or an observation into the sample is influenced by unobserved variables  (e.g., the respondent's level of trust toward healthcare entities may cause them to self-select out of the study or refuse to sign consent for a test), which in turn are correlated with unobserved variables related to an outcome of interest (e.g., the result of a blood test). [@vella1998] 
\begin{figure}[h!]
\vspace{-0.7cm}
   \centering
   \includegraphics[scale=0.35]{DAG_heckman.png}
   \caption{DAG Heckman selection model: here the nodes (dotted lines = latent variables, continuous lines = observed variables) describe the relationship between $y^*_{ij}$ the latent response and $r^*_{ij}$ the latent selection variables of a $j$th unit in a $i$th cluster,that are dependable on $x_{ij}^O$ and $x_{ij}^S$ set of predictors and are correlated through $u_{ij}$ unobservable variables.}
   \label{fig:DAG}
\end{figure}
This is visualized in Figure \ref{fig:DAG}, where for the $j$th individual or unit within the $i$th cluster, there is $y_{ij}^*$, a latent outcome variable, and $r_{ij}^*$, a latent selection variable, which are correlated through $u_{ij}$ an unobserved or unrecorded variable, with $i\in[1,2,...,N]$ and $j\in[1,2,...,n_i]$. Here, both latent variables are related to the sets of predictor covariates $x_{ij}^O$ and $x_{ij}^S$. From $r_{ij}^*$ one can derive $r_{ij}=I(r_{ij}^*>=0)$ a selection indicator of $y_{ij}^*$ into the sample, and with this in turn, one can define $y_{ij}=y_{ij}^*,\forall r_{ij}=1$ the observable outcome variable.

@@

Generally the same variables can be used on the matrix of predictor variables $X^O_{i}$ and $X^S_{i}$. However, to avoid multicollinearity problems, [@puhani1997] it is recommended to include in $X^S_{i}$ at least one variable that is not included in the outcome model. [@angrist2001] This variable is commonly known as an exclusion restriction variable (ERV), and should only be associated with the selection in the sample $r^*_{ij}$ (relevance condition) but not with the actual observation $y^*_{ij}$ (exclusion condition).[@gomes_etal20] The ERV should meet the relevance and exclusion conditions in order to provide independent information about the selection processing in order to facilitate the estimation of the Heckman model specially when dataset does not follow the BVN assumption.
<!-- Can you give some explanation (1-3 sentences?) on the exclusion and restriction assumptions here? -->

In the presence of selection bias, aforementioned outcome equation will yield biased estimates of $\beta^O_{i}$ if no efforts are made to adjust for the non-representativeness of the observed $X^O_{i}$ and $y_{i}$ values. For this reason, the Heckman model aims to jointly estimate the outcome and selection equation by defining a relation between their respective error distributions. For instance, Heckman’s original model [@heckman1976] assumes that residual terms have a bivariate normal distribution (BVN),

@@

The interpretation of $\rho_i$ is fairly straightforward. When $\rho_i=0$, the participation does not affect the outcome model and missing data can be considered MCAR (if data are missing completely at random) or MAR (if missingness is already explained by $x_{ij}^O$). Conversely, when $\rho_i\neq 0$, this suggests that data are MNAR.

@@

Under the BVN assumption, the parameters of the Heckman model coefficients can be estimated using the two-step Heckman method (H2S) [@heckman1976] or the full information maximum likelihood method (FIML).[@amemiya1984] However, both methods can lead to inconsistent estimators when this underlying distribution cannot be assumed. [@gomes_etal20] To overcome this problem, other approaches have been proposed that relax the distribution assumptions, among them copula models. [@smith03] The copula approach uses a function, known as a copula, that joins the marginal distribution of the error terms of the selection and outcome equation which are specified separately.
$$ F(r^*_{i},y^*_{i}) =\Phi\left(r^*_{i}-X^S_{i}\beta^S_{i},\frac{y^*_{i}-X^O_{i}\beta^O_{i}}{\sigma_i}, \rho_i\right)$$
Thus, to estimate the parameters of the Heckman method, it is sufficient to specify the marginal distributions of the error terms, and link them with a suitable copula function. In our imputation method we estimate the Heckman model using the copula method, as in real life many data could not follow a BVN distribution. This facilitates the Heckman model estimation and makes it more robust to deviations from the assumptions regarding the distribution of the data.

@@

The Heckman model can be extended to hierarchical settings, i.e., in data where individuals or sampling units are nested within groups, as is the case in EHR or IPD. In this case, sample units from the same group are expected to share some characteristics (e.g., distribution of variables, relationships between exposure and outcomes, missing processes, missing mechanisms) given only the fact that they belong to the same group.

Since most statistical analyses assume that the sample units are independent of each other, more complex hierarchical models dealing with nested data are required.  This hierarchical complexity must not only be taken into account in the observation process, but also in the missing data process, thus requiring imputation models that are congenial to the analysis model, i.e., that have the same assumptions about the data.

Different procedures can be adopted to combine information between groups; however, in our imputation method we opted for the two-stage approach that is often used in meta-analyses. This is because such an approach is less computationally intensive and could potentially generate fewer convergence problems in the estimation of the Heckman hierarchical model compared to other approaches.

<!-- This sentence was too long and has the verbs all the way at the end, making it hard to read. I suggest to rephrase to three shorter sentences: OK!!-->
Briefly, in a first stage  we estimate $\theta_i$ the cluster specific parameters of the Heckman model. That is, we estimate separately for each of the $N$ clusters,the parameters of the two equations in each study, the outcome model and the selection model. In the second stage, we combine all $\theta_i$  by using a random effects meta-analysis model.

<!-- I wouldn't call the distribution imaginary. It is assumed to be real but latent. Some words we could use: conjectured, latent, presupposed, presumed, Thanks :D -->
In a random effects model, $\theta_i$ parameters are assumed to be drawn independently and identically from a latent distribution of parameters with a  population mean $\theta_m$ and a population variance $\psi$. [@higgins2009] Thus, the  $\theta_i=\theta_m + b_i$ can be specified using $b_i \sim N(0,\psi)$ random effects to allow for between-study heterogeneity in observed data relationships, and between-study heterogeneity in missing patterns.

# Method
\label{sec:methods}
We follow a similar approach proposed by Resche-Rigon and White (2018) [@resche-rigon2018] for multilevel imputation of data. Briefly, their method was developed to impute variables from a hierarchical structure (i.e., when there are samples unit grouped within a cluster or group).
This method involves estimating an outcome equation (describing the relationships of the observed data to the missing variable) separately in each cluster, after which the parameter estimates of that equation are pooled using random effects meta-analysis.

With this method, values can be imputed in very common scenarios in IPD, e.g., sporadic and systematic missing patterns. In particular, when the response variable is systematically missing within a group, i.e., when $y_{ij}$ are totally missing within a group, the imputation values are drawn from a (generalized) linear model conditional on $\theta_m$ the marginal population parameters, i.e., those estimated after pooling the cluster-specific parameters $\theta_i$. On the other hand, when the variable is sporadically missing within the group, i.e., there are some observed $y_{ij}$ within the cluster, the imputation model is conditional on the shrunk-cluster parameters, i.e., those coming from the shrinkage of $\theta_i$ towards $\theta_m$.

Our imputation approach differs crucially from the previous approach as here we estimate two correlated equations (instead of a single outcome equation) in each cluster, thus obtaining $\theta_i=\{\beta_i^O,\beta_i^S,\sigma_i,\rho_i\}$ parameters from both equations which are then pooled into a $\theta_m$ parameter set at the marginal or population level. Our imputation method was created for impute univariate incomplete datasets, but since it is implemented in a Gibbs sampling procedure, it can also be used to impute multiple incomplete variables in a data set.
<!-- I don't understand what you mean here: it is univariate and it can be used to impute multiple variables? I would say these two statements contradict each other. yes you are right i tried to say is a mehtod for univariate incomplete dataset -->

## Imputation of univariate incomplete dataset
Given an outcome variable $y=(y_1,y_2,...,y_N)^T$, that consists of $y^{miss}_{ij}$ missing and $y^{obs}_{ij}$ observable values, we generate independent draws from the posterior predictive distribution for the missing data, $y_{ij}^{miss}$, given the observable data information $y_{ij}^{obs}$.
$$p(y_{ij}^{miss}|y_{ij}^{obs})=\int_{\theta} p(y_{ij}^{miss}|\theta,y_{ij}^{obs})p(\theta|y_{ij}^{obs})\mathrm{d}\theta$$
Here we implicitly assume vague prior distributions for each of the parameters included in the parameter vector $\theta$.  Because the integration can be performed computationally by sampling from the posterior predictive distribution $p(\theta|y_{ij}^{obs})$, our imputation method can be carried out in the following two steps:

1. Draw a ${\theta}$ parameter vector, ${\theta^*}$, from $p(\theta|y_{ij}^{obs})$, their posterior distribution.

2. Draw $y_{ij}^{miss}$ from $p(y_{ij}^{miss}|\theta^*)$, their predictive distribution for a given $\theta^*$ vector.

@@

### Draw the $\theta^*$ parameter vector

#### Fit $p(y_{ij}^{obs}|\theta_i)$, the heckman selection model at group level

Initially, we use the copula method to estimate the set of cluster-specific parameters, $\widehat{\theta_i}=\{\widehat{\beta^O_i}, \widehat{\beta^S_i}, \widehat{\sigma_i}, \widehat{\rho_i}\}$, using all $j$ units with observable measurements $y_{ij}^{obs}$ within each group $i$. 
The Heckman model is estimated with the \textbf{gjrm} function of the GJRM R package, [@radiceGJRMGeneralisedJoint2021] under the bivariate model with the nonrandom sample selection (BSS) specification, from which we obtain not only the parameters' point estimates $\widehat{\theta_i}$, but also their corresponding $\widehat{S(\theta_i)}$ within-cluster variance-covariance matrix.

<!-- Can you add a reference to the R package here? OK -->

<!-- When the model does not converge in a study, we assume that data follows a MAR mechanism so we fixed $\widehat{\rho_i} = 0$ and we estimate only the outcome equation parameters from a (generalized) linear model. -->

#### Fit a meta-analysis model
<!-- This paragraph starts of in the the present tense " we pool ", and later it's past tense " we pooled" etc. Stick to one tense. In applied papers, it is normal to use only past tense. In stats papers it is common to use present tense when describing a (new) method in the methods section. -->
In this step, we pool the parameters $\widehat{\theta_i}$ with a random effects meta-analysis model using only the groups with observable information, i.e., those that are not systematically missing and have sufficient information to estimate the heckman model.  In particular, we pool the $p$ coefficients of the $\beta^O$ outcome equation and estimated a multivariate random effects meta-analysis model with them, similarly we combine all $q$ coefficient parameters of the $\beta^S$ selection equation. 
<!-- The phrase " the log-transformed parameter of $\sigma$" is a description of $\sigma'$, so it is what is known as a commenting or parenthetical clause. That means you need to put a comma before and after this clause: "... on $\sigma'$, the log-transformed parameter of $\sigma$, ...". Although this article is not for the BMJ, I can recommend reading their house-style for writing: https://www.bmj.com/about-bmj/resources-authors/house-style -->
<!-- Read more here if you like: http://content.principia.edu/teaching-excellence/commenting-with-commas-when-and-where-to-use-them/ Thanks a lot you are a walking library :D, very appreciated all your lessons!! -->
We also performed a univariate random effects meta_analysis on $\sigma'$, the log-transformed parameter of $\sigma$, and another on $\rho'$, the fisher-transformed parameter of $\rho$.

The meta-analysis model is performed with the \textbf{mixmeta} function of the R package mixmeta[@gasparrini2021], which allows the use of maximum likelihood (ML), restricted maximum likelihood (REML), and moments estimation methods. For the simulation and illustrative study, we used the restricted REML estimation method, which is recommended as it has a good balance between unbiasedness and efficiency. [@viechtbauer2005]
<!-- Insensitivity is not generally a good thing. I think you should either explain why it is a good thing here: "... it is insensitive to the true value of the mean of the distribution of random effects and it is efficient" (or something similar, feel free to rephrase) , or simplify it to "... a good balance between unbiasedness and efficiency. jajaj I think at certain point i used the translator to correct my writing and i did not notice that change unbiasedness by insensitive..But thanks God you have eagle eyes @@) .."  -->

#### Draw $\theta^*_m$ the marginal parameters

From the meta-analysis model, we obtain the marginal estimates $\widehat{\theta_m}$ and the between-cluster variance $\widehat{\psi}$ with their corresponding variance-covariance matrices $\widehat{S_{\theta_m}}$ and $\widehat{S_{\psi}}$, which are used to draw the $\theta^*_m$ and $\psi^*$ parameters as follows:

\begin{align*}
\theta_m^*& \sim N(\widehat{\theta_m},\widehat{S_{\theta_m}})\\
\psi^*& \sim N(\widehat{\psi},\widehat{S_{\psi}})
\end{align*}

#### Draw the cluster parameters $\theta^*_i$
<!-- Throughout the text, consider the order in which you put the symbol and the text: -->
<!-- Put the symbol after the text if the text is a description of the symbol: "shrunk-cluster parameters $\theta^*_i$" . -->
<!-- Put the text after the symbol if the symbol quantifies the text: "$n$ patients" , or " $N$ clusters" . Whoahh you are a gramatical Lord thanks a lot!! -->
We draw the shrunk-cluster parameters $\theta^*_i$ for each group $i$ from the following posterior distribution conditional on $\theta^*_m$ and $\psi^*$.
\begin{align*}
\theta^*_i\sim N\left(\frac{ \theta^*_m/\psi^*+\widehat{\theta_i}/\widehat{S_{\theta_i}}}{1/\psi^*+1/\widehat{S_{\theta_i}}},\frac{1}{1/\psi^*+1/\widehat{S_{\theta_i}}}\right)
\end{align*}

As can be seen, the mean and variance of the posterior distribution is a combination between the estimated marginal and cluster-specific parameters. Here the weights on the cluster-specific parameters $\widehat{\theta_i}$ and the marginal parameters $\theta^*_m$ are inversely proportional to the within cluster variance $\widehat{S_{\theta_i}}$ and between clusters variance $\psi^*$. For example, when $\widehat{S_{\theta_i}} < \psi^*$ the mean of the conditional distribution gives more weight to the estimated cluster-specific parameter.
Conversely, when $\widehat{S_{\theta_i}}>\psi^*$, more weight is given to the estimated marginal parameters. Therefore, in case of a cluster with systematic missingness, it is as if the within-cluster variance is infinite ($\widehat{S_{\theta_i}}\to\infty$), so all the weight is assigned to the parameters estimated at the marginal level. That is, when there is no information in a cluster, we rely entirely on the marginal information.

<!-- What does then mean in this last part? should it be " and lettting all parameters rely ..." ? I think now is more clear, thanks!! -->

@@

Having $\theta^*_i$, the shrunk-cluster parameters vector for each group, we back-transform $\sigma^*$ and $\rho^*$ to the original scale. Then $y^{miss}_{ij}$, the missing values, can be drawn from $p(y^{miss}_{ij}|\theta^*_i)$, their predictive distribution given $\theta^*_i$, as follows:

@@

The imputed value of $y_{ij}^{miss}$ can be drawn from the conditional expectation of $y_{ij}$ on unobserved measurements:
\begin{align*}
\mu&= E[y_{ij}|r_{ij}=0,{\beta^O_{i}}^*,{\beta^{S}_{i}}^*,{\rho_{i}}^*,{\sigma_i}^*]\\
\mu&= x^O_{ij}{\beta^O_{i}}^*+\rho_{i}^*\sigma_i^*\frac{-\phi(x^{S}_{ij}{\beta^{S}_{i}}^*)}{\Phi(-x^{S}_{ij}{\beta^{S}_i}^*)}\\
y_{ij}^{miss}&\sim N(\mu,{\sigma_i^*}^2)
\end{align*}

@@

<!-- The next paragraph belongs in the manual for the package, or a vignette. I don't think it is necessary to put it here. 
The MICE package allows one to specify imputation methods to each of the missing variables by setting the method and the predictive matrix for each of the missing variables. To use our proposed method, it is necessary to specify for the MNAR missing variable the method \textbf{2l.heckman} in the mice methods vector. Furthermore, in the prediction matrix, the group or cluster variable should be specified as \textbf{'-2'}, all predictor variables belonging to the selection and outcome as \textbf{'1'}, the exclusion restrictions or predictor variables that are only included in the selection equation as \textbf{'-3'} and those that are only included in the outcome equation as \textbf{'-4'}. Please refer to the toy example in the attached github repository to better understand how to implement the imputation model.   -->

@@

## Aim
We designed a simulation study aimed to compare the performance of imputation methods for imputing a missing variable in a hierarchical dataset, where the missingness follows a MNAR mechanism.

@@

To investigate the performance of the imputation methods under the following scenarios:

>>>>

corres: "* Johanna Munoz. \\email{j.munozavila@umcutrecht.nl}"
presentaddress: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Universiteitsweg 100, 3584 CG Utrecht, the Netherlands
authormark: Uthor \emph{et al}.
articletype: Research article
received: 2022-11-01
revised: 2022-12-01
accepted: 2023-01-01
abstract: "Missing data is a common problem in medical research, and is commonly addressed using multiple imputation. Although traditional imputation methods allow for valid statistical inference when data are missing at random (MAR), their implementation is problematic when the presence of missingness depends on unobserved information, i.e. the data are missing not at random (MNAR). Unfortunately, this situation is rather common, in observational studies, registries and other sources of real-world data. While several imputation methods have been proposed for addressing individual studies when data are MNAR, their application and validity in large datasets with clustering remains unclear. We therefore explored the consequence of MNAR data in hierarchical data in-depth, and proposed a novel multilevel imputation method for common missing patterns in clustered datasets. This method is based on the principles of Heckman selection models and adopts a two-stage meta-analysis approach to impute binary and continuous variables that may be outcomes or predictors and that are systematically or sporadically missing. After evaluating the proposed imputation model in simulated scenarios, we illustrate it use in a cross-sectional community survey to estimate the prevalence of malaria parasitemia in children aged 2-10 years in five subregions in Uganda."

@@

Selection bias occurs when the inclusion of a subject or an observation into the sample is influenced by unobserved variables  (e.g., the respondent's level of trust toward healthcare entities may cause them to self-select out of the study or refuse to sign consent for a test), which in turn are correlated with unobserved variables related to an outcome of interest (e.g., the result of a blood test). [@vella1998] 
\begin{figure}[h!]
\vspace{-0.7cm}
   \centering
   \includegraphics[scale=0.35]{DAG_heckman.png}
   \caption{DAG Heckman selection model: here the nodes (dotted lines = latent variables, continuous lines = observed variables) describe the relationship between $y^*_{ij}$ the latent response and $r^*_{ij}$ the latent selection variables of a $j$th unit in a $i$th cluster,that are dependable on $ \boldsymbol{x_{ij}^O}$ and $ \boldsymbol{x_{ij}^S}$ set of predictors and are correlated through $ \boldsymbol{u_{ij}}$ unobservable variables.}
   \label{fig:DAG}
\end{figure}
This is visualized in Figure \ref{fig:DAG}, where for the $j$th individual or unit within the $i$th cluster, there is $y_{ij}^*$, a latent outcome variable, and $r_{ij}^*$, a latent selection variable, which are correlated through $\boldsymbol{u_{ij}}$ an unobserved or unrecorded variable, with $i\in[1,2,...,N]$ and $j\in[1,2,...,n_i]$. Here, both latent variables are related to the sets of predictor covariates $\boldsymbol{x_{ij}^O}$ and $\boldsymbol{x_{ij}^S}$. From $r_{ij}^*$ one can derive $r_{ij}=I(r_{ij}^*>=0)$ a selection indicator of $y_{ij}^*$ into the sample, and with this in turn, one can define $y_{ij}=y_{ij}^*,\forall r_{ij}=1$ the observable outcome variable.

@@

Generally the same variables can be used on the matrix of predictor variables $\boldsymbol{X^O_{i}}$ and $\boldsymbol{X^S_{i}}$. However, to avoid multicollinearity problems, [@puhani1997] it is recommended to include in $\boldsymbol{X^S_{i}}$ at least one variable that is not included in the outcome model. [@angrist2001] This variable is commonly known as an exclusion restriction variable (ERV), and should only be associated with the selection in the sample $r^*_{ij}$ (relevance condition) but not with the actual observation $y^*_{ij}$ (exclusion condition).[@gomes_etal20] The ERV should meet the relevance and exclusion conditions in order to provide independent information about the selection process that facilitates the estimation of the Heckman model.

In the presence of selection bias,the aforementioned outcome equation will yield biased estimates of $\boldsymbol{\beta^O_{i}}$ if no efforts are made to adjust for the non-representativeness of the observed $\boldsymbol{X^O_{i}}$ and $\boldsymbol{y_{i}}$ values. For this reason, the Heckman model aims to jointly estimate the outcome and selection equation by defining a relation between their respective error distributions. For instance, Heckman’s original model [@heckman1976] assumes that residual terms have a bivariate normal distribution (BVN),

@@

The interpretation of $\rho_i$ is fairly straightforward. When $\rho_i=0$, the participation does not affect the outcome model and missing data can be considered MCAR (if data are missing completely at random) or MAR (if missingness is already explained by $\boldsymbol{x_{ij}^O}$). Conversely, when $\rho_i\neq 0$, this suggests that data are MNAR.

@@

Under the BVN assumption, the parameters of the Heckman model coefficients can be estimated using the two-step Heckman method (H2S) [@heckman1976] or the full information maximum likelihood method (FIML).[@amemiya1984] However, both methods may lead to inconsistent estimators when the true underlying distribution is not the BVN. [@gomes_etal20] To overcome this problem, other approaches have been proposed that relax the distribution assumptions, among them copula models. [@smith03] The copula approach uses a function, known as a copula, that joins the marginal distribution of the error terms of the selection and outcome equation which are specified separately. The Heckman model can be expressed with the following bivariate normal copula:
$$ F(\boldsymbol{r^*_{i}},\boldsymbol{y^*_{i}}) =\Phi\left(\boldsymbol{r^*_{i}}-\boldsymbol{X^S_{i}}\boldsymbol{\beta^S_{i}},\frac{\boldsymbol{y^*_{i}}-\boldsymbol{X^O_{i}}\boldsymbol{\beta^O_{i}}}{\sigma_i}, \rho_i\right),$$
where $\Phi$ is the bivariate standard normal cumulative distribution function.
Thus, to estimate the parameters of the Heckman method, it is sufficient to specify the marginal distributions of the error terms, and link them with a suitable copula function. In our imputation method we estimate the Heckman model using the copula method, as in real life data generating mechanism may deviate a BVN distribution. This facilitates the Heckman model estimation and makes it more robust to deviations of distributional assumptions.

@@

The Heckman model can be extended to hierarchical settings, i.e., in cases when individuals or sampling units are nested within clusters, as is the case in EHR or IPD. In this case, sample units from the same cluster are expected to share some characteristics (e.g., the distribution of variables and correlations between them, relationships between exposure and outcomes, missing processes).

Since most statistical analyses assume that the sample units are independent of each other, more complex hierarchical models dealing with nested data are required.  This hierarchical complexity must not only be taken into account in the observation process,i.e. model that connects outcome and predictors, but also in the missing process,i.e related to the distribution and reasons for missingness, thus requiring imputation models that are congenial to the analysis model, i.e., that make the same assumptions about the data.

Different procedures can be adopted to combine information between clusters; however, in our imputation method we opted for the two-stage approach that is often used in meta-analyses.[@simmonds2005] This is because such an approach is computationally less intensive and could potentially generate fewer convergence problems in the estimation of the Heckman hierarchical model compared to other approaches.

Briefly, in a first stage  we estimate $\boldsymbol{\theta_i}$ the cluster specific parameters of the Heckman model. That is, we estimate separately for each of the $N$ clusters,the parameters of the two equations in each study, the outcome model and the selection model. In the second stage, we combine all $\boldsymbol{\theta_i}$  by using a random effects multivariate meta-analysis model.

In a random effects model, $\boldsymbol{\theta_i}$ parameters are assumed to be drawn independently and identically from a latent multivariate normal distribution of parameters with a  population mean $\boldsymbol{\theta_m}$ and a population variance $\boldsymbol{\psi}$. [@higgins2009] Thus, we can write $\boldsymbol{\theta_i}=\boldsymbol{\theta_m} + \boldsymbol{b_i}$  using $\boldsymbol{b_i} \sim N(0,\boldsymbol{\psi})$ random effects to allow for between-study heterogeneity in observed data relationships, and between-study heterogeneity in missing patterns.

# Using the Heckman model to impute missing data
\label{sec:methods}
We follow a similar approach proposed by Resche-Rigon and White (2018) [@resche-rigon2018] for multilevel imputation of data. Briefly, their method was developed to impute variables from a hierarchical structure (i.e., when there are units grouped within a cluster or group).
This method involves estimating an outcome equation (describing the relationships of the observed data to the missing variable) separately in each cluster, after which the parameter estimates of that equation are pooled using random effects meta-analysis.

With this method, values can be imputed in very common scenarios in IPD, e.g., sporadic and systematic missing patterns. In particular, when the response variable is systematically missing within a group, i.e., when $y_{ij}$ are totally missing within a cluster, the imputation values are drawn from a (generalized) linear model conditional on $\boldsymbol{\theta_m}$ the marginal population parameters, i.e., those estimated after pooling the cluster-specific parameters $\boldsymbol{\theta_i}$. On the other hand, when the variable is sporadically missing within the cluster, i.e., there are some observed $y_{ij}$ within the cluster, the imputation model is conditional on the shrunk-cluster parameters, i.e., those coming from the shrinkage of $\boldsymbol{\theta_i}$ towards $\boldsymbol{\theta_m}$.

Our imputation approach differs crucially from the previous approach as here we estimate two correlated equations (instead of a single outcome equation) in each cluster, thus obtaining $\boldsymbol{\theta_i}=\{\boldsymbol{\beta_i^O},\boldsymbol{\beta_i^S},\sigma_i,\rho_i\}$ parameters from both equations which are then pooled into a $\boldsymbol{\theta_m}$ parameter set at the marginal or population level. Our imputation method was created for imputing datasets with a single missing outcome or covariate variable, but it can also be used to impute multiple incomplete variables in a data set.

In the following subsections, we explain the method when the incomplete variable is the outcome variable but this method can also be used to impute any incomplete MNAR predictor variable in the dataset. In the latter case, in the outcome equation (\ref{eq:1}), the incomplete predictor variable must be specified as the dependent variable and $\boldsymbol{X^O_{i}}$ as the set of covariates associated with it. On the other hand, in the selection model(\ref{eq:2}), $\boldsymbol{r^*_{i}}$ corresponds to the latent variable of selection of the incomplete predictor variable together with its $\boldsymbol{X^S_{i}}$ associated set of predictors.

## Univariate imputation
Given an outcome variable $\boldsymbol{y}=(y_1,y_2,...,y_N)^T$, that consists of $y^{miss}_{ij}$ missing and $y^{obs}_{ij}$ observable values, we generate independent draws from the posterior predictive distribution for the missing data, $y_{ij}^{miss}$, given the observable data information $y_{ij}^{obs}$.
$$p(y_{ij}^{miss}|y_{ij}^{obs})=\int_{\boldsymbol{\theta}} p(y_{ij}^{miss}|\boldsymbol{\theta},y_{ij}^{obs})p(\boldsymbol{\theta}|y_{ij}^{obs})\mathrm{d}\boldsymbol{\theta}$$
Here we implicitly assume vague prior distributions for each of the parameters included in the parameter vector $\boldsymbol{\theta}$.  Because the integration can be performed computationally by sampling from the posterior predictive distribution $p(\boldsymbol{\theta}|y_{ij}^{obs})$, our imputation method can be carried out in the following two steps:

1. Draw a $\boldsymbol{\theta}$ parameter vector, $\boldsymbol{\theta^*}$, from $p(\boldsymbol{\theta}|y_{ij}^{obs})$, their posterior distribution.

2. Draw $y_{ij}^{miss}$ from $p(y_{ij}^{miss}|\boldsymbol{\theta^*})$, their predictive distribution for a given $\boldsymbol{\theta^*}$ vector.

@@

### Draw the $\boldsymbol{\theta^*}$ parameter vector

#### Fit $p(y_{ij}^{obs}|\boldsymbol{\theta_i})$, the heckman selection model at cluster level

Initially, we use the copula method to estimate the set of cluster-specific parameters, $\boldsymbol{\widehat{\theta_i}}=\boldsymbol{\{\widehat{\beta^O_i}}, \boldsymbol{\widehat{\beta^S_i}}, \widehat{\sigma_i}, \widehat{\rho_i}\}$, using all $j$ units with observable measurements $y_{ij}^{obs}$ within each cluster $i$. 
The Heckman model is estimated with the \textbf{gjrm} function of the GJRM R package, [@radiceGJRMGeneralisedJoint2021] under the bivariate model with the nonrandom sample selection (BSS) specification, from which we obtain not only the parameters' point estimates $\boldsymbol{\widehat{\theta_i}}$, but also their corresponding $\boldsymbol{\widehat{S(\theta_i)}}$ within-cluster variance-covariance matrix.

#### Fit a meta-analysis model

In this step, we pool the parameters $\boldsymbol{\widehat{\theta_i}}$ with a random effects meta-analysis model using only the clusters with observable information, i.e., those with no systematically missing outcome. In particular, we pool the $p$ coefficients of the $\boldsymbol{\beta^O}$ in the outcome equation and estimate a multivariate random effects meta-analysis model with them. Similarly we combine all $q$ coefficient parameters of the $\boldsymbol{\beta^S}$ in the selection equation.

We also perform a univariate random effects meta_analysis on $\sigma'$, the log-transformed parameter of $\sigma$, and another on $\rho'$, the fisher-transformed parameter of $\rho$.

The meta-analysis model is performed with the \textbf{mixmeta} function of the R package mixmeta[@gasparrini2021], which allows the use of maximum likelihood (ML), restricted maximum likelihood (REML), and moments estimation methods. For the simulation and illustrative study, we use the restricted REML estimation method, which is recommended as it has a good balance between unbiasedness and efficiency. [@viechtbauer2005]

#### Draw $\boldsymbol{\theta^*_m}$ the marginal parameters

From the meta-analysis model, we obtain the marginal estimates $\boldsymbol{\widehat{\theta_m}}$ and the between-cluster variance matrix $\boldsymbol{\widehat{\psi}}$,i.e. variance-covariance matrix of the random effects, with their corresponding variance-covariance matrices $\boldsymbol{\widehat{S_{\theta_m}}}$ and $\boldsymbol{\widehat{S_{\psi}}}$, which are used to draw the $\boldsymbol{\theta^*_m}$ and $\boldsymbol{\psi^*}$ parameters, from their posterior distribution as follows: [@resche-rigon2013]

\begin{align*}
\boldsymbol{\theta_m^*}& \sim N(\boldsymbol{\widehat{\theta_m}},\boldsymbol{\widehat{S_{\theta_m}}})\\
\boldsymbol{\psi^*}& \sim N(\boldsymbol{\widehat{\psi}},\boldsymbol{\widehat{S_{\psi}}})
\end{align*}

#### Draw the cluster parameters $\boldsymbol{\theta^*_i}$

We draw the shrunked-cluster-parameters $\boldsymbol{\theta^*_i}$ for each cluster $i$ from the following posterior distribution conditional on $\boldsymbol{\theta^*_m}$ and $\boldsymbol{\psi^*}$.

\begin{align*}
\theta^*_i\sim N\left(\left({\boldsymbol{\psi^*}}^{-1}+{\boldsymbol{\widehat{S_{\theta_i}}}}^{-1}\right)^{-1}\left(\boldsymbol{\psi^*}^{-1}\boldsymbol{\theta^*_m}+\boldsymbol{\widehat{S_{\theta_i}}}^{-1}\boldsymbol{\widehat{\theta_i}} \right),\left({\boldsymbol{\psi^*}}^{-1}+{\boldsymbol{\widehat{S_{\theta_i}}}}^{-1}\right)^{-1} \right)
\end{align*}

As can be seen, the mean and variance of the posterior distribution is a combination between the estimated marginal and cluster-specific parameters. Here the weights on the cluster-specific parameters $\boldsymbol{\widehat{\theta_i}}$ and the marginal parameters $\boldsymbol{\theta^*_m}$ are inversely proportional to the within cluster variance $\boldsymbol{\widehat{S_{\theta_i}}}$ and between clusters variance $\boldsymbol{\psi^*}$. For example, when $\boldsymbol{\widehat{S_{\theta_i}}} < \boldsymbol{\psi^*}$ the mean of the conditional distribution gives more weight to the estimated cluster-specific parameter.
Conversely, when $\boldsymbol{\widehat{S_{\theta_i}}}>\boldsymbol{\psi^*}$, more weight is given to the estimated marginal parameters. Therefore, in case of a cluster with systematic missingness, it is as if the within-cluster variance is infinite ($\boldsymbol{\widehat{S_{\theta_i}}}\to\infty$), so all the weight is assigned to the parameters estimated at the marginal level. That is, when there is no information in a cluster, we rely entirely on the marginal information.

@@

Having $\boldsymbol{\theta^*_i}$, the shrunk-cluster parameters vector for each cluster, we back-transform $\sigma^*$ and $\rho^*$ to the original scale. Then $y^{miss}_{ij}$, the missing values, can be drawn from $p(y^{miss}_{ij}|\boldsymbol{\theta^*_i})$, their predictive distribution given $\boldsymbol{\theta^*_i}$, as follows:

@@

The imputed value of $y_{ij}^{miss}$ can be drawn from the conditional expectation of $y_{ij}$ on unobserved measurements: [@greene2018]
\begin{align*}
\mu&= E[y_{ij}|r_{ij}=0,{\boldsymbol{\beta^O_{i}}^*},\boldsymbol{{\beta^{S}_{i}}^*},{\rho_{i}}^*,{\sigma_i}^*]\\
\mu&= \boldsymbol{x^O_{ij}}\boldsymbol{{\beta^O_{i}}^*}+\rho_{i}^*\sigma_i^*\frac{-\phi(\boldsymbol{x^{S}_{ij}}\boldsymbol{{\beta^{S}_{i}}^*})}{\Phi(-\boldsymbol{x^{S}_{ij}}{\boldsymbol{\beta^{S}_i}^*})}\\
y_{ij}^{miss}&\sim N(\mu,{\sigma_i^*}^2)
\end{align*}

where $\phi(.)$ is the probability density function and $\Phi(.)$ is the cumulative distribution function of the standard normal distribution.

@@

@@

## Aim
We designed a simulation study aimed to compare the performance of alternative methods for imputing a single missing outcome variable in a hierarchical dataset, where the missingness follows a MNAR mechanism. In our scenarios we considered systematically missingness,i.e. the case some of the studies did not report hte variable at all.

@@

In addition to the basic scenario described above, we explored additional data generating mechanisms. Specifically we investigated the performance of the imputation methods under the following scenarios:
